---
title: "Faulkner and punctuation"
author: "Johannes Burgers"
format: html
editor: visual
---

## Introduction

There is a lot to say about Faulkner and punctuation.

In this pieces I will argue: Faulkner pushes the very definition of a sentence in order to demonstrate the continuity of time. Yet, how he experiments with punctuation to achieve this end changes? throughout his career.

## Critical literature on Faulkner


## Methodology and Editing Faulkner


## Computation

```{r packages}
library(tidyverse)

```

```{r load_data}
all_works <-
  list.files(path = "./data/", pattern = "*.txt") %>% #grab a list of all the files with .txt extension
  map_df(~ data_frame(  #the map function performs the same command on all parts of the data set. In this case the .txt files
    text = read_file("./data/"), #read the files
    date = ifelse(
      str_detect(basename(.), "[:digit:]{4}") == TRUE,
      str_extract(basename(.), "[:digit:]{4}"),
      NA
    ) #see if there is a date in yyyy format, i.e. 1987, and extract the date, if it can't find date NA
  )) 

```

